---
title: Batch correction
output:
  html_document:
    toc: true
---

Last updated: `r as.character(Sys.Date())`

Code version: `r system("git log -1 --format=oneline | cut -d' ' -f1", intern = TRUE)`

Data version: `r system("git --git-dir=../data/.git log -1 --format=oneline | cut -d' ' -f1", intern = TRUE)`

```{r setup, echo = FALSE}
source("knitr_options.R")
opts_chunk$set(message = FALSE)
```

## Executive summary

*  Simply regressing using the base R function `lm` without protecting the variables of interest is effective.
*  The function `removeBatchEffect` from the limma package and `ComBat` from the sva package peform very similarly.
*  `removeBatchEffect` is more flexible than `ComBat` and thus is preferred in situations in which there is more than one batch effect or if a variable to regress is continuous.
*  The functions `sva` and `svaseq` both failed to capture the clear batch effect present in this data set.

## Introduction

I perform the following methods to remove batch effects and compare their first two principal components.

*  **raw**: No manipulation
*  **lm**: Simply regress the batch effects using R's `lm` function
*  **limma**: Use `removeBatchEffect` to regress batch effects while protecting variables of interest.
*  **combat**: Use `ComBat` in two steps to regress batch effects without protecting variables of interest
*  **combat_var**: Use `ComBat` in two steps to regress batch effects while protecting variables of interest
*  **sva**: Use `sva` to regress unspecified batch effects.
*  **svaseq**: Use `sva` to regress unspecified batch effects from the count data.

## Prepare data

```{r packages}
library("limma")
library("edgeR")
library("sva")
library("ggplot2")
theme_set(theme_bw(base_size = 16))
```

Read in the data of counts (row = sample, column = feature).

```{r data}
full <- read.table("../data/counts_per_sample.txt", header = TRUE,
                   stringsAsFactors = FALSE)
full <- full[order(full$dir), ]
rownames(full) <- paste(full$dir, full$ind, full$bact, full$time, sep = ".")
counts <- t(full[, grep("ENSG", colnames(full))])
# Filter lowly expressed genes
counts <- counts[rowSums(cpm(counts) > 1) > 6, ]
```

The first `r sum(!grepl("ENSG", colnames(full)))` columns are meta-data about each sample, and the remaining are the gene counts.
Only the `r nrow(counts)` genes which have at least one read per million (pre-normalization with TMM) in at least six of the `r nrow(full)` samples are kept.

Convert the meta-data to factors for modeling.

```{r annotation}
groups <- full[, c("ind", "bact", "time", "extr", "rin")]
groups$bact <- gsub("\\+", "plus", groups$bact)
groups$ind <- factor(groups$ind)
groups$bact <- factor(groups$bact, levels = c("none", "Rv", "Rvplus", "GC",
                                              "BCG", "Smeg", "Yers", "Salm",
                                              "Staph"))
groups$time <- factor(groups$time, levels = c(4, 18, 48))
groups$extr <- factor(groups$extr)
head(groups)
```

Normalize with TMM (`calcNormFactors`).

```{r model}
y <- DGEList(counts)
y <- calcNormFactors(y)
```

Create a matrix of expression values (log~2~ counts per million) for visualizing the data.
Since `y` is an object of class DGElist, the TMM normalized library sizes are automatically used.
Also, a prior count of 0.25 is added to avoid taking the log of zero.

```{r log-cpm}
dat_cpm <- cpm(y, log = TRUE)
# Remove leading number from sample names. I had given them the unique 
# identifieers 1-156, which were useful when pre-processing the data but are no 
# longer needed.
colnames(dat_cpm) <- sapply(strsplit(colnames(dat_cpm), "\\."),
                            function(x) {paste(x[2:4], collapse = ".")})
```

## Functions

```{r pca}
run_pca <- function(exp_mat) {
  # Runs principal components analysis.
  # Input:
  #  exp_mat: data frame or matrix of expression values.
  #           Rows are genes and samples are columns.

  # Including some default settings just to be explicit
  pca <- prcomp(x = t(exp_mat), retx = TRUE, center = TRUE, scale. = TRUE)
  return(pca)
}
```

```{r plot-pca}
plot_pca <- function(pca, anno, title) {
  # Plot PC1 vs PC2.
  # Input:
  #  pca: result from prcomp
  #  anno: data frame that contains bact and time columns
  #  title: string to use as title for plot
  stopifnot(nrow(pca$x) == nrow(anno))
  pca_anno <- cbind(anno, pca$x)
  pca_anno$control <- ifelse(pca_anno$bact == "none", "control", "infection")
  pca_anno$control <- factor(pca_anno$control, levels = c("control",
                                                          "infection"))
  ggplot(pca_anno, mapping = aes(x = PC1, y = PC2,
                                 col = time, size = control)) +
    geom_point() +
    scale_size_discrete(range = c(5, 2)) +
    scale_color_brewer(palette = "Dark2") +
    labs(title = title)
}
```

## Raw data

PCA of uncorrected log~2~ transformed cpm values:

```{r pca-uncorrected}
pca_raw <- run_pca(dat_cpm)
(pca_plot_raw <- plot_pca(pca_raw, groups, title = "PCA: Uncorrected log CPM expression matrix"))
```

## lm from base R

Remove effect of RIN score and extraction batch.

```{r regress-lm}
dat_lm <- dat_cpm
for (i in 1:nrow(dat_lm)) {
  lm_model <- lm(dat_cpm[i, ] ~ extr + rin, data = groups)
  resids <- residuals(lm_model)
  dat_lm[i, ] <- resids + lm_model$coefficients[1] # Add back intercept term
}
```

PCA of batch-corrected log~2~ transformed cpm values:

```{r pca-lm}
pca_lm <- run_pca(dat_lm)
(pca_plot_lm <- plot_pca(pca_lm, groups,
                         title = "PCA: Regress with lm"))

```

## removeBatchEffect from limma

Remove effect of RIN score and extraction batch.

```{r regress}
# Regress out the effect of RIN and extraction batch
design_base <- model.matrix(~ind + bact*time, data = groups)
design_base <- design_base[, -ncol(design_base)] # Remove Staph.48
dat_limma <- removeBatchEffect(x = dat_cpm, batch = groups$extr,
                               covariates = groups$rin,
                               design = design_base)
```

PCA of batch-corrected log~2~ transformed cpm values:

```{r pca-corrected}
pca_limma <- run_pca(dat_limma)
(pca_plot_limma <- plot_pca(pca_limma, groups,
                            title = "PCA: removeBatchEffect"))

```

## ComBat without variable of interest

ComBat is ideal when there is only one categorical variable to correct.
If you have more than one batch to correct, you either need to combine the factors into one large factor (if the combined factor has at least 3-4 samples per level) or perform ComBat correction multiple times ([source][combat-multiple-batch]).
It is not possible to correct for a continuous variable.

[combat-multiple-batch]: https://support.bioconductor.org/p/57894/

I convert the RIN score into a categorical variable with 5 levels.

```{r categorical-rin}
summary(groups$rin)
groups$rin_cat <- cut(groups$rin, 5)
summary(groups$rin_cat)
```

There are not enough samples per level to create a combined factor, so I need to perform ComBat twice.

```{r}
table(groups$extr, groups$rin_cat)
```

First I correct for extraction batch, while protecting RIN.

```{r combat-extr}
modcombat <- model.matrix(~rin_cat, data = groups)
dat_combat_1 <- ComBat(dat = dat_cpm,
                     batch = groups[, "extr"],
                     mod = modcombat,
                     par.prior = TRUE, prior.plots = FALSE)
```

Second I correct for RIN.

```{r combat-rin}
modcombat <- model.matrix(~1, data = groups)
dat_combat_2 <- ComBat(dat = dat_combat_1,
                       batch = groups[, "rin_cat"],
                       mod = modcombat,
                       par.prior = TRUE, prior.plots = FALSE)
```

```{r}
pca_combat <- run_pca(dat_combat_2)
(pca_plot_combat <- plot_pca(pca_combat, groups,
                            title = "PCA: ComBat without variable of interest"))
```

## ComBat with variable of interest

Whether or not to protect the variable(s) of interest when using ComBat is currently a work-in-progress.
Initially it was advised to include the variable(s) of interest (as reflected in the [ComBat function documentation][combat-doc]).
However, more recently the advice has been changed to not include the variable(s) of interest (as reflected in the [manual][sva-man]).
See [here](https://support.bioconductor.org/p/63082/), [here](https://support.bioconductor.org/p/63007/), and [here](https://support.bioconductor.org/p/62874/#62875) for discussions.

[combat-doc]: http://www.rdocumentation.org/packages/sva/functions/ComBat
[sva-man]: http://www.bioconductor.org/packages/release/bioc/vignettes/sva/inst/doc/sva.pdf

First I correct for extraction batch, while protecting RIN and the variables of interest: individual, bacteria, time,  and bacteria-time interactions.

```{r combat-extr-var-of-interest}
modcombat <- model.matrix(~rin_cat + ind + bact*time, data = groups)
modcombat <- modcombat[, -ncol(modcombat)] # Remove Staph.48
dat_combat_var_1 <- ComBat(dat = dat_cpm,
                     batch = groups[, "extr"],
                     mod = modcombat,
                     par.prior = TRUE, prior.plots = FALSE)
```

Second I correct for RIN while still protecting variables of interest.

```{r combat-rin-var-of-interest}
modcombat <- model.matrix(~ind + bact*time, data = groups)
modcombat <- modcombat[, -ncol(modcombat)] # Remove Staph.48
dat_combat_var_2 <- ComBat(dat = dat_combat_var_1,
                       batch = groups[, "rin_cat"],
                       mod = modcombat,
                       par.prior = TRUE, prior.plots = FALSE)
```

```{r}
pca_combat_var <- run_pca(dat_combat_var_2)
(pca_plot_combat_var <- plot_pca(pca_combat_var, groups,
                            title = "PCA: ComBat with variable of interest"))
```

## sva

Predict the number of surrogate variables.

```{r}
mod0 <- model.matrix(~1, data = groups)
(n_sv = num.sv(dat_cpm, design_base, method = "leek"))
```

Run sva.

```{r sva}
svobj = sva(dat_cpm, design_base, mod0, n.sv = n_sv)
```

The result didn't make any sense.
It just assigned the first sample 1 and all the other samples 0.

```{r}
table(svobj$sv)
plot(svobj$sv)
```

So not surprisingly this doesn't really change anything.

```{r}
dat_sva <- removeBatchEffect(x = dat_cpm,
                             covariates = svobj$sv,
                             design = design_base)
```

```{r}
pca_sva <- run_pca(dat_sva)
(pca_plot_combat_var <- plot_pca(pca_sva, groups,
                            title = "PCA: sva"))
```

## svaseq

There is also a method that can be applied directly to the count data.
It performed just as poorly as the standard `sva` function.

```{r}
svseqobj = svaseq(counts, design_base, mod0, n.sv = n_sv)
```

```{r}
table(svseqobj$sv)
plot(svseqobj$sv)
```

## Comparing methods

Compare PC1 and PC2 from all the methods.

```{r}
results <- data.frame(raw_pc1 = pca_raw$x[, 1],
                      raw_pc2 = pca_raw$x[, 2],
                      lm_pc1 = pca_lm$x[, 1],
                      lm_pc2 = pca_lm$x[, 2],
                      limma_pc1 = pca_limma$x[, 1],
                      limma_pc2 = pca_limma$x[, 2],
                      combat_pc1 = pca_combat$x[, 1],
                      combat_pc2 = pca_combat$x[, 2],
                      combat_var_pc1 = pca_combat_var$x[, 1],
                      combat_var_pc2 = pca_combat_var$x[, 2])
summary(results)
```

```{r corr-table, results='asis'}
kable(cor(results), format = "markdown", digits = 2)
```

```{r heatmap, fig.height=10}
library("gplots")
heatmap.2(as.matrix(results), trace = "none", margins = c(10, 5))
cor(results$limma_pc1, results$combat_pc1)
```

The results from all the correction methods are highly correlated.
The results "limma" and "combat_var" cluster together because they both protect the variables of interest.
The results "lm" and "combat" cluster together because they both do **not** protect the variables of interest.

## Session information

```{r info}
sessionInfo()
```
