Hierarchical clustering
==========================================================

John Blischak

Last updated: `r as.character(Sys.Date())`

Code version: `r system("git log -1 --format=oneline | cut -d' ' -f1", intern = TRUE)`

Data version: `r system("git --git-dir=../data/.git log -1 --format=oneline | cut -d' ' -f1", intern = TRUE)`

Heatmaps displaying hierarchical clustering of samples.
The distance metric is Euclidean distance (the default for `dist`).
The clustering method is complete linkage (the default for `hclust`).

**Outline:**

+ All expressed** genes
  + Median expression levels for each bacteria:time combination (n = 6)
  + Per bacterial infection
  + Per timepoint
  + Per individual
+ Immune-related genes
  + Median expression levels for each bacteria:time combination (n = 6)
  + Per bacterial infection
  + Per timepoint
  + Per individual
  
** "expressed" genes defined as having greater than one read per million in at least 6 out of the 156 samples

```{r global-chunk-options, echo = FALSE}
source("knitr_options.R")
opts_chunk$set(echo = FALSE, warning = FALSE, fig.width = 12, fig.height = 12)
```

```{r packages, message = FALSE}
library(ggplot2)
library(gridExtra)
library(heatmap.plus)
library(colorspace)
library(rols)
library(biomaRt)
library(edgeR)
library(dplyr)
```

```{r obtain-immune-genes}
go_im <- olsQuery('immune', 'GO')
go_im_names <- names(go_im)

ensembl <- useMart('ensembl', dataset = 'hsapiens_gene_ensembl')
g_im <- getBM(attributes = c('ensembl_gene_id'),
           filters = c('go_id'),
           values = list(go_im_names),
           mart = ensembl)
```

```{r functions}
custom_heatmap <- function(mat, dist_metric = "euclidean", ...) {
  # mat: numeric matrix, row = sample, col = gene
  # dist_metric: method to use for `dist`
  # ... additional parameters passed to `heatmap.plus`, e.g. ColSideColors
  d <- dist(mat, method = dist_metric)
  heatmap.plus(as.matrix(d), margins = c(10, 10),
               cexRow = 1.5, cexCol = 1.5, ...)
}
```

```{r input-data}
load("../data/model-fit.rda")
normalized_lib_size <- fit$samples$lib.size * fit$samples$norm.factors
dat_cpm <- cpm(fit$counts, lib.size = normalized_lib_size)
bact_vec <- c("Rv", "GC", "Rv+", "BCG", "Smeg", "Salm", "Yers", "Staph")
# bact_vec <- c("BCG", "Rv", "Staph")  # for debugging
anno <- read.table("../data/annotation.txt", header = TRUE,
                   stringsAsFactors = TRUE)
# Convert time to a factor
anno$time <- ordered(anno$time, levels = c(4, 18, 48))
cpm_per_sample <- cbind(anno, t(dat_cpm))
# Remove sample number from rownames
names_split <- strsplit(rownames(cpm_per_sample), "\\.")
rownames(cpm_per_sample) <- sapply(names_split, function(x)
                                                paste(x[2:4], collapse = "."))
```

# Expressed genes

## Median expression levels

```{r medians}
# Group on multiple variables with unknown names
group_vars <- c("bact", "time")
group_vars <- lapply(group_vars, as.symbol)
cpm_grouped <- regroup(cpm_per_sample, group_vars)
# Calculate median expression for each of the gene columns
genes <- grep("ENSG", colnames(cpm_per_sample), value = TRUE)
dots <- lapply(genes, function(x) substitute(median(x), list(x=as.name(x))))
cpm_medians <- do.call(summarise, c(list(.data = cpm_grouped), dots))
# Fix the gene column names
colnames(cpm_medians)[grep("ENSG", colnames(cpm_medians))] <- genes
write.table(cpm_medians, file = "../data/median_expression.txt",
            quote = FALSE, sep = "\t", row.names = FALSE)
```

```{r median-heatmap}
genes <- cpm_medians[, grep("ENSG", colnames(cpm_medians))]
rownames(genes) <- paste(as.character(cpm_medians$bact),
                         as.character(cpm_medians$time), sep = ".")
color_bact <- cpm_medians$bact
levels(color_bact) = rainbow_hcl(length(levels(color_bact)))
color_time <- cpm_medians$time
levels(color_time) = c("green", "blue", "purple")
color_col_median <- matrix(c(as.character(color_time),
                             as.character(color_bact)), ncol = 2)
colnames(color_col_median) <- c("time", "bact")
custom_heatmap(genes, ColSideColors = color_col_median)
legend("topleft", c(levels(cpm_medians$bact), levels(cpm_medians$time)), 
       fill = c(levels(color_bact), levels(color_time)),
       title = "Legend")
```

---

## Per bacterial infection

```{r bacteria-expr, results = "asis"}
for (b in bact_vec) {
  cat("\n\n### ", b, "\n\n")
  b_sub <- cpm_per_sample[cpm_per_sample$bact %in% c("none", b), ]
  b_sub <- droplevels(b_sub)
  b_genes <- b_sub[, grep("ENSG", colnames(b_sub))]
  # Color factors
  color_bact <- b_sub$bact
  if (levels(color_bact)[1] == "none") {
    levels(color_bact) = c("black", "red")
  } else {
    levels(color_bact) = c("red", "black")
  }
  color_time <- b_sub$time
  levels(color_time) = c("green", "blue", "purple")
  color_ind <- b_sub$ind
  levels(color_ind) = rainbow_hcl(length(levels(color_ind)))
  color_col <- matrix(c(as.character(color_ind), as.character(color_time),
                        as.character(color_bact)), ncol = 3)
  colnames(color_col) <- c("ind", "time", "bact")
  custom_heatmap(b_genes, ColSideColors = color_col)
  legend("topleft",
         c(levels(b_sub$bact), levels(b_sub$time), levels(b_sub$ind)), 
         fill = c(levels(color_bact), levels(color_time), levels(color_ind)),
         title = "Legend")
}
```

---

## Per timepoint

```{r time-expr, results = "asis"}
for (tp in levels(cpm_per_sample$time)) {
  cat("\n\n### ", tp, "\n\n")
  tp_sub <- cpm_per_sample[cpm_per_sample$time == tp, ]
  tp_sub <- droplevels(tp_sub)
  tp_genes <- tp_sub[, grep("ENSG", colnames(tp_sub))]
  # Color factors
  color_bact <- tp_sub$bact
  levels(color_bact) = rainbow_hcl(length(levels(color_bact)))
  color_ind <- tp_sub$ind
  levels(color_ind) = rainbow_hcl(length(levels(color_ind)))
  color_col <- matrix(c(as.character(color_ind), as.character(color_bact)),
                      ncol = 2)
  colnames(color_col) <- c("ind", "bact")
  custom_heatmap(tp_genes, ColSideColors = color_col)
  legend("topleft", c(levels(tp_sub$bact), levels(tp_sub$ind)), 
         fill = c(levels(color_bact), levels(color_ind)),
         title = "Legend")
}
```

---

## Per individual

```{r ind-expr, results = "asis"}
for (ind in levels(cpm_per_sample$ind)) {
  cat("\n\n### ", ind, "\n\n")
  ind_sub <- cpm_per_sample[cpm_per_sample$ind == ind, ]
  ind_sub <- droplevels(ind_sub)
  ind_genes <- ind_sub[, grep("ENSG", colnames(ind_sub))]
  # Color factors
  color_bact <- ind_sub$bact
  levels(color_bact) = rainbow_hcl(length(levels(color_bact)))
  color_time <- ind_sub$time
  levels(color_time) = c("green", "blue", "purple")
  color_col <- matrix(c(as.character(color_time), as.character(color_bact)),
                      ncol = 2)
  colnames(color_col) <- c("time", "bact")
  custom_heatmap(ind_genes, ColSideColors = color_col)
  legend("topleft", c(levels(tp_sub$bact), levels(tp_sub$time)), 
         fill = c(levels(color_bact), levels(color_time)),
         title = "Legend")
}
```

---

---

# Immune-related genes

## Median expression levels

```{r median-heatmap-immune}
genes_im <- genes[, colnames(genes) %in% g_im$ensembl_gene_id]
custom_heatmap(genes_im, ColSideColors = color_col_median)
legend("topleft", c(levels(cpm_medians$bact), levels(cpm_medians$time)), 
       fill = c(levels(color_bact), levels(color_time)),
       title = "Legend")
```

---

## Per bacterial infection

```{r bacteria-im, results = "asis"}
for (b in bact_vec) {
  cat("\n\n### ", b, "\n\n")
  b_sub <- cpm_per_sample[cpm_per_sample$bact %in% c("none", b), ]
  b_sub <- droplevels(b_sub)
  b_genes <- b_sub[, grep("ENSG", colnames(b_sub))]
  b_genes_im <- b_genes[, colnames(b_genes) %in% g_im$ensembl_gene_id]
  # Color factors
  color_bact <- b_sub$bact
  if (levels(color_bact)[1] == "none") {
    levels(color_bact) = c("black", "red")
  } else {
    levels(color_bact) = c("red", "black")
  }
  color_time <- b_sub$time
  levels(color_time) = c("green", "blue", "purple")
  color_ind <- b_sub$ind
  levels(color_ind) = rainbow_hcl(length(levels(color_ind)), c = 150, l = 50)
  color_col <- matrix(c(as.character(color_ind), as.character(color_time),
                        as.character(color_bact)), ncol = 3)
  colnames(color_col) <- c("ind", "time", "bact")
  custom_heatmap(b_genes_im, ColSideColors = color_col)
  legend("topleft",
         c(levels(b_sub$bact), levels(b_sub$time), levels(b_sub$ind)), 
         fill = c(levels(color_bact), levels(color_time), levels(color_ind)),
         title = "Legend")
}
```

---

## Per timepoint

```{r time-im, results = "asis"}
for (tp in levels(cpm_per_sample$time)) {
  cat("\n\n### ", tp, "\n\n")
  tp_sub <- cpm_per_sample[cpm_per_sample$time == tp, ]
  tp_sub <- droplevels(tp_sub)
  tp_genes <- tp_sub[, grep("ENSG", colnames(tp_sub))]
  tp_genes_im <- tp_genes[, colnames(tp_genes) %in% g_im$ensembl_gene_id]
  # Color factors
  color_bact <- tp_sub$bact
  levels(color_bact) = rainbow_hcl(length(levels(color_bact)))
  color_ind <- tp_sub$ind
  levels(color_ind) = rainbow_hcl(length(levels(color_ind)))
  color_col <- matrix(c(as.character(color_ind), as.character(color_bact)),
                      ncol = 2)
  colnames(color_col) <- c("ind", "bact")
  custom_heatmap(tp_genes_im, ColSideColors = color_col)
  legend("topleft", c(levels(tp_sub$bact), levels(tp_sub$ind)), 
         fill = c(levels(color_bact), levels(color_ind)),
         title = "Legend")
}
```

---

## Per individual

```{r ind-im, results = "asis"}
for (ind in levels(cpm_per_sample$ind)) {
  cat("\n\n### ", ind, "\n\n")
  ind_sub <- cpm_per_sample[cpm_per_sample$ind == ind, ]
  ind_sub <- droplevels(ind_sub)
  ind_genes <- ind_sub[, grep("ENSG", colnames(ind_sub))]
  ind_genes_im <- ind_genes[, colnames(ind_genes) %in% g_im$ensembl_gene_id]
  # Color factors
  color_bact <- ind_sub$bact
  levels(color_bact) = rainbow_hcl(length(levels(color_bact)))
  color_time <- ind_sub$time
  levels(color_time) = c("green", "blue", "purple")
  color_col <- matrix(c(as.character(color_time), as.character(color_bact)),
                      ncol = 2)
  colnames(color_col) <- c("time", "bact")
  custom_heatmap(ind_genes_im, ColSideColors = color_col)
}
```

```{r info}
sessionInfo()
```
