Identify genes in GO categories
========================================================

John Blischak

Last updated: `r as.character(Sys.Date())`

Code version: `r system("git log -1 --format=oneline | cut -d' ' -f1", intern = TRUE)`

Data version: `r system("git --git-dir=../data/.git log -1 --format=oneline | cut -d' ' -f1", intern = TRUE)`

```{r global-chunk-options, echo = FALSE}
source("knitr_options.R")
opts_chunk$set(fig.width = 12, fig.height = 8, cache = FALSE)
```

```{r libraries, message = FALSE}
library("topGO")
```

## Load data

Import the motif results.

```{r}
table_s3 <- read.table("../data/table-s3.txt", header = TRUE, sep = "\t",
                       stringsAsFactors = FALSE)
```

Import the GO results.

```{r}
go_18_all <- read.delim("../data/go-18-k5-All.txt",
                        stringsAsFactors = FALSE)
go_18_mtb <- read.delim("../data/go-18-k5-MTB.txt",
                        stringsAsFactors = FALSE)
go_18_non <- read.delim("../data/go-18-k5-non-DE.txt",
                        stringsAsFactors = FALSE)
go_18_vir <- read.delim("../data/go-18-k5-Virulent.txt",
                        stringsAsFactors = FALSE)
go_18_yer <- read.delim("../data/go-18-k5-Yers-Salm.txt",
                        stringsAsFactors = FALSE)
go_48_all <- read.delim("../data/go-48-k5-All.txt",
                        stringsAsFactors = FALSE)
go_48_mtb <- read.delim("../data/go-48-k5-MTB.txt",
                        stringsAsFactors = FALSE)
go_48_non <- read.delim("../data/go-48-k5-non-DE.txt",
                        stringsAsFactors = FALSE)
go_48_vir <- read.delim("../data/go-48-k5-Virulent.txt",
                        stringsAsFactors = FALSE)
go_48_yer <- read.delim("../data/go-48-k5-Yers-Salm.txt",
                        stringsAsFactors = FALSE)
go_all_18_48 <- read.delim("../data/go-all-k6-18 & 48 h.txt",
                        stringsAsFactors = FALSE)
go_all_18 <- read.delim("../data/go-all-k6-18 h.txt",
                        stringsAsFactors = FALSE)
go_all_48 <- read.delim("../data/go-all-k6-48 h.txt",
                        stringsAsFactors = FALSE)
go_all_all <- read.delim("../data/go-all-k6-All.txt",
                        stringsAsFactors = FALSE)
go_all_non <- read.delim("../data/go-all-k6-non-DE.txt",
                        stringsAsFactors = FALSE)
go_all_yer <- read.delim("../data/go-all-k6-Yers-Salm.txt",
                        stringsAsFactors = FALSE)
```

## Create GO data

Create a topGO object with fake DE data.
This does not matter because only need the available universe of genes.
I get the significance results from the text files loaded above.
The GO analysis is too slow to run repeatedly.

```{r create-go-data}
gene_universe <- rep(c(0, 1), each = nrow(table_s3) / 2)
gene_universe <- factor(gene_universe)
names(gene_universe) <- table_s3$id
go_data <- new("topGOdata",
               ontology = "BP",
               allGenes = gene_universe,
               nodeSize = 5,
               annotationFun = annFUN.org,
               mapping = "org.Hs.eg",
               ID = "ensembl")
```

## Function for finding intersection

```{r}
find_genes <- function(go_id, m_genes, m_names, go_data) {
  # go_id: GO ID
  # m_genes: list of candidate Ensembl gene IDs
  # m_names: list of names corresponding to m_genes
  # go_data: the topGO object which contains the GO database mappings
  go_genes <- genesInTerm(go_data, go_id)[[1]]
  total_go <- length(go_genes)
  total_m <- length(m_genes)
  hits <- which(m_genes %in% go_genes)
  cat(sprintf("%d genes are shared between the %d candidate genes and the %d genes in the GO category",
              length(hits), total_m, total_go))
  result <- data.frame(id = m_genes[hits],
                       name = m_names[hits],
                       stringsAsFactors = FALSE)
  return(result)
}
```

## Fig2 - Full time course

### All

```{r}
go_all_all[grep("GO:0060337", go_all_all$GO.ID), ]
interferon_2 <- find_genes(go_id = "GO:0060337",
                  m_genes = table_s3$id[table_s3$full_time_course == "All"],
                  m_names = table_s3$name[table_s3$full_time_course == "All"],
                  go_data = go_data)
cat(sort(interferon_2$name))
go_all_all[grep("GO:0002739", go_all_all$GO.ID), ]
cytokine_2 <- find_genes(go_id = "GO:0002739",
                  m_genes = table_s3$id[table_s3$full_time_course == "All"],
                  m_names = table_s3$name[table_s3$full_time_course == "All"],
                  go_data = go_data)
cat(sort(cytokine_2$name))
go_all_all[grep("GO:2000109", go_all_all$GO.ID), ]
apop_3 <- find_genes(go_id = "GO:2000109",
                  m_genes = table_s3$id[table_s3$full_time_course == "All"],
                  m_names = table_s3$name[table_s3$full_time_course == "All"],
                  go_data = go_data)
cat(sort(apop_3$name))
go_all_all[grep("GO:0008630", go_all_all$GO.ID), ]
apop_4 <- find_genes(go_id = "GO:0008630",
                  m_genes = table_s3$id[table_s3$full_time_course == "All"],
                  m_names = table_s3$name[table_s3$full_time_course == "All"],
                  go_data = go_data)
cat(sort(apop_4$name))
```

### 18 & 48 h

```{r}
go_all_18_48[grep("GO:0050766", go_all_18_48$GO.ID), ]
phago <- find_genes(go_id = "GO:0050766",
                  m_genes = table_s3$id[table_s3$full_time_course == "18 & 48 h"],
                  m_names = table_s3$name[table_s3$full_time_course == "18 & 48 h"],
                  go_data = go_data)
cat(sort(phago$name))
```

## Fig3 - 18 hour analysis

### MTB

```{r}
go_18_mtb[grep("monocyte", go_18_mtb$Term), ]
mtb_mono <- find_genes(go_id = "GO:0045657",
                  m_genes = table_s3$id[table_s3$time_18h == "MTB"],
                  m_names = table_s3$name[table_s3$time_18h == "MTB"],
                  go_data = go_data)
cat(sort(mtb_mono$name))
```

```{r}
go_18_mtb[grep("vesicle", go_18_mtb$Term), ]
mtb_vesicle <- find_genes(go_id = "GO:0031338",
                  m_genes = table_s3$id[table_s3$time_18h == "MTB"],
                  m_names = table_s3$name[table_s3$time_18h == "MTB"],
                  go_data = go_data)
cat(sort(mtb_vesicle$name))
```

### Virulent

```{r}
go_18_vir[grep("GTPase", go_18_vir$Term), ]
vir_gtp <- find_genes(go_id = "GO:0051057",
                  m_genes = table_s3$id[table_s3$time_18h == "Virulent"],
                  m_names = table_s3$name[table_s3$time_18h == "Virulent"],
                  go_data = go_data)
cat(sort(vir_gtp$name))
```


## Fig4 - 48 hour analysis

### MTB

```{r}
go_48_mtb[grep("itamin", go_48_mtb$Term), ]
vit_d <- find_genes(go_id = "GO:0033280",
                  m_genes = table_s3$id[table_s3$time_48h == "MTB"],
                  m_names = table_s3$name[table_s3$time_48h == "MTB"],
                  go_data = go_data)
cat(sort(vit_d$name))
```

AQP3 is an aquaporin.

CXCL10 is also known as Interferon gamma-induced protein 10.

PIM1 is involved in cytokine signaling.

http://www.biomedcentral.com/1471-2199/13/18/

TNC is a glycoprotein regulated by growth factors.

### Virulent

```{r}
go_48_vir[grep("macrophage", go_48_vir$Term), ]
mac_activation <- find_genes(go_id = "GO:0043032",
                  m_genes = table_s3$id[table_s3$time_48h == "Virulent"],
                  m_names = table_s3$name[table_s3$time_48h == "Virulent"],
                  go_data = go_data)
cat(sort(mac_activation$name))
```

DE in cattle in response to _M. bovis_

http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0041066#s3

Another DE study in mononuclear cells.

http://www.sciencedirect.com/science/article/pii/S016158900700795X

```{r}
go_48_vir[grep("antigen", go_48_vir$Term), ]
vir_antigen <- find_genes(go_id = "GO:0002580",
                  m_genes = table_s3$id[table_s3$time_48h == "Virulent"],
                  m_names = table_s3$name[table_s3$time_48h == "Virulent"],
                  go_data = go_data)
cat(sort(vir_antigen$name))
```


```{r info}
sessionInfo()
```
