Boxplots for phagosome maturation
========================================================

Last updated: `r as.character(Sys.Date())`

Code version: `r system("git log -1 --format=oneline | cut -d' ' -f1", intern = TRUE)`

Data version: `r system("git --git-dir=../data/.git log -1 --format=oneline | cut -d' ' -f1", intern = TRUE)`

Plotting boxplots of genes for phagosome maturation at 18 hours for Figure S4.

```{r}
library("ggplot2")
library("reshape2")
```

Load Table S3 just to get the gene name conversions.

```{r input}
table_s3 <- read.table("../data/table-s3.txt", header = TRUE, sep = "\t",
                       stringsAsFactors = FALSE)
anno <- read.table("../data/annotation.txt", header = TRUE,
                   stringsAsFactors = FALSE)
```

```{r}
gene_conver <- table_s3$id
names(gene_conver) <- table_s3$name
```

```{r}
target_name <- c("RAB22A", "RAP2A", "RAB3A", "RAB33A", "ATP6V1D")
target_id <- gene_conver[target_name]
target_time <- 18
```

Read in log fold changes, where each sample was compared to its time-matched control (the two values were subtracted).

```{r}
dat_fc <- read.table("../data/fold-change.txt", header = TRUE,
                     stringsAsFactors = FALSE)
dat_fc <- dat_fc[dat_fc$time %in% target_time,
                 c("ind", "bact", "time", target_id)]
stopifnot(colnames(dat_fc)[4:(3 + length(target_id))] == target_id)
colnames(dat_fc)[4:(3 + length(target_id))] <- target_name
head(dat_fc)
```

```{r}
dat_fc$bact <- ordered(dat_fc$bact, levels = c("Rv", "Rv+", "GC", "BCG",
                                               "Smeg", "Yers", "Salm", "Staph"))
```

```{r}
library("RColorBrewer")
my_cols <- brewer.pal(n = 9, name = "Greens")
my_cols <- c(rev(my_cols)[1:5], "purple", "blue", "orange")
```

Melt to long form.

```{r}
dat_fc_long <- melt(dat_fc, id.vars = c("ind", "bact", "time"), variable.name = "gene",
                    value.name = "log2FC")
head(dat_fc_long)
```

Plot.

```{r plot-function}
phago <- ggplot(dat_fc_long,  aes(x = bact, y = log2FC, fill = bact)) +
  geom_boxplot(lwd = 1) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  facet_wrap(~gene, nrow = 4, scales = "free") +
  labs(x = "Bacterial infection",
       y = expression(paste(log[2], " fold change"))) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(size = rel(1.25)),
        line = element_line(size = 1), rect = element_rect(size = 1),
        legend.position="none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_manual(values = my_cols)
phago
```

```{r}
pdf("../figs/fig-S04-phago.pdf", width = 8, height = 7,
    useDingbats = FALSE)
phago
dev.off()
```


```{r info}
sessionInfo()
```
