Plot over time
========================================================

Last updated: `r as.character(Sys.Date())`

Code version: `r system("git log -1 --format=oneline | cut -d' ' -f1", intern = TRUE)`

Data version: `r system("git --git-dir=../data/.git log -1 --format=oneline | cut -d' ' -f1", intern = TRUE)`

Plot log2 fold change over time for full joint analysis (Fig 2B).

```{r global-chunk-options, echo = FALSE}
source("knitr_options.R")
opts_chunk$set(fig.width = 12, fig.height = 12, warning = FALSE)
```

```{r}
library("ggplot2")
library("plyr")
library("reshape2")
```

Log2 fold change measurements are stored in `fold-change.txt`.

```{r input}
dat_fc <- read.table("../data/fold-change.txt", header = TRUE,
                     stringsAsFactors = FALSE)
dat_fc[, c("dir", "extr", "rin", "mol", "index", "master.mix")] <- list(NULL)
motifs <- read.table("../data/table-s3.txt", header = TRUE, sep = "\t",
                     stringsAsFactors = FALSE)
```

Get the mean log2 fold change per bacterial infection (mean across individuals).

```{r}
dat_fc <- ddply(dat_fc, c("bact", "time"), colwise(mean, is.numeric))
```

Take the absolute value.

```{r}
dat_fc[, -1:-2] <- abs(dat_fc[, -1:-2])
```

Melt, add motifs, and average across bacterial infections.

```{r}
dat_melt <- melt(dat_fc, id.vars = c("bact", "time"), variable.name = "id",
                 value.name = "log2fc")
dat_motif <- merge(dat_melt, motifs, by = "id")
dat_mean <- ddply(dat_motif, c("time", "bact", "full_time_course"), colwise(mean, is.numeric))
```

Order factors

```{r}
dat_mean$bact <- ordered(dat_mean$bact, levels = c("Rv", "Rv+", "GC", "BCG",
                                               "Smeg", "Yers", "Salm", "Staph"))
dat_mean$full_time_course <- ordered(dat_mean$full_time_course,
                                     levels = c("non-DE", "All", "18 & 48 h",
                                                "48 h", "18 h", "Yers-Salm"))
```

Color scheme

```{r}
library("RColorBrewer")
my_cols <- brewer.pal(n = 9, name = "Greens")
my_cols <- c(rev(my_cols)[1:5], "purple", "blue", "orange")
```

Plot

```{r}
plot_over_time <- ggplot(dat_mean, aes(x = time, y = log2fc, col = bact)) +
  geom_line() + geom_point() +
  facet_wrap(~ full_time_course, nrow = 2) +
  labs(x = "Hours post-infection",
       y = expression(paste("Mean ", log[2],
                            " fold change in expression (absolute value)")),
       col = "Bacteria") +
  theme_bw(base_size = 18) + scale_x_continuous(breaks = c(4, 18, 48)) +
  scale_colour_manual(values = my_cols) +
  scale_y_continuous(limits = c(-.25, 3), breaks = 0:3, labels = 0:3) +
  theme(legend.title = element_text(size = rel(1)),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank())
plot_over_time
```

```{r}
pdf("../figs/fig-02B-plot-over-time.pdf", width = 10, height = 7,
    useDingbats = FALSE)
plot_over_time
dev.off()
```

```{r info}
sessionInfo()
```
