Process supplemental data from PNAS paper
========================================================

John Blischak

Last updated: `r as.character(Sys.Date())`

Code version: `r system("git log -1 --format=oneline | cut -d' ' -f1", intern = TRUE)`

Data version: `r system("git --git-dir=../data/.git log -1 --format=oneline | cut -d' ' -f1", intern = TRUE)`

Process supplemental data from PNAS paper.
[Barreiro et al., 2012][barreiro2012] infected dendritic cells (DCs) with MTB and identified response eQTLs.

[barreiro2012]: http://www.pnas.org/content/109/4/1204.long

```{r global-chunk-options, echo = FALSE}
source("knitr_options.R")
```

Download and process DC expression data (Table S1).

```{r s01-expression}
library("gdata")
dc_expr <- read.xls("http://www.pnas.org/content/suppl/2012/01/10/1115761109.DCSupplemental/sd01.xls",
                    skip = 1, nrow = 12958, stringsAsFactors = FALSE)
rownames(dc_expr) <- dc_expr$Ensembl.gene.ID
dc_expr <- dc_expr[, c(-1, -6)]
colnames(dc_expr) <- c("name", "logFC", "PValue", "FDR")
write.table(dc_expr, file = "../data/dc-expr.txt", quote = FALSE, sep = "\t")
head(dc_expr)
```

Download and process DC eQTL data (Table S3).

```{r s03-eQTL}
dc_eqtl <- read.xls("http://www.pnas.org/content/suppl/2012/01/10/1115761109.DCSupplemental/sd03.xls",
                 skip = 2, nrow = 11954, stringsAsFactors = FALSE)
write.table(dc_eqtl, file = "../data/dc-eqtl.txt", quote = FALSE, sep = "\t")
dc_eqtl[1, ]
```

Classify DC eQTL results into categories:

*  "response": eQTL pre- **or** post-infection
*  "general": eQTL pre- **and** post-infection
*  "non": eQTL pre- **nor** post-infection

```{r}
eqtl_type <- character(length = nrow(dc_eqtl))
for (i in 1:nrow(dc_eqtl)) {
  if (dc_eqtl$General.eQTL_2f[i] == "YES") {
    eqtl_type[i] <- "general"
  } else if (dc_eqtl$NI.Only.eQTL_2f[i] == "YES" | 
             dc_eqtl$MTB.Only.eQTL_2f[i] == "YES") {
    eqtl_type[i] <- "response"
  } else {
    eqtl_type[i] <- "non"
  }
}
eqtl_type <- data.frame(eqtl_type, row.names = dc_eqtl$Ensembl.gene.ID,
                        stringsAsFactors = FALSE)
table(eqtl_type$eqtl_type)
```

Combine eQTL classifications with differential expression results from DCs.

```{r}
expr_eqtl <- merge(dc_expr, eqtl_type, by = "row.names")
colnames(expr_eqtl)[1] <- "id"
stopifnot(nrow(expr_eqtl) <= min(nrow(dc_expr), nrow(eqtl_type)))
stopifnot(nrow(expr_eqtl) > 0)
write.table(expr_eqtl, file = "../data/dc-expr-eqtl.txt", quote = FALSE, sep = "\t",
            row.names = FALSE)
head(expr_eqtl)
```

```{r info}
sessionInfo()
```
