Data from Tailleux et al., 2008
========================================================

Last updated: `r as.character(Sys.Date())`

Code version: `r system("git log -1 --format=oneline | cut -d' ' -f1", intern = TRUE)`

Data version: `r system("git --git-dir=../data/.git log -1 --format=oneline | cut -d' ' -f1", intern = TRUE)`

```{r setup, echo = FALSE}
source("knitr_options.R")
opts_chunk$set(message = FALSE, fig.width = 10, fig.height = 8)
```

[Tailleux et al., 2008][Tailleux2008] studied the transcriptional response to MTB infection in human macrophages and dendritic cells.
Transcriptome data was also collected from MTB.
The human data was deposited in the GENOPOLIS databases (www.genopolis.it), but this resource no longer exists.
Ludovic had given me the data.
It contains a file with the normalized data, `exprSetSnorm.txt`, a report from the [AMDA][] software, `report.pdf`, and a bunch of other files that I do not use below.

In this analysis, I perform the following:

*  Confirm that the data is normalized
*  Filter the genes based on expression level
*  Perform PCA

[Tailleux2008]: http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0001403
[AMDA]: http://www.biomedcentral.com/1471-2105/7/335

## Setup

```{r packages}
library("ggplot2")
theme_set(theme_bw(base_size = 16))
library("gridExtra")
library("biomaRt")
library("dplyr")
```

Input the data.

```{r data}
exp_norm <- read.table("../data/exprSetSnorm.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
stopifnot(ncol(exp_norm) == 72)
exp_norm[1:10, 1:5]
```

Log2 transform the data.

```{r log2}
exp_norm <- log2(exp_norm)
```

There are `r ncol(exp_norm)` samples:
9 donors x 2 cell types x 4 time points.

Update the sample names to follow the naming convention I use.

```{r update-names}
samples <- colnames(exp_norm)
cell_type <- substr(samples, 1, 1)
time <- substr(samples, 2, 3)
time <- sub("h", "", time)
time <- as.numeric(time)
ind <- strsplit(samples, "h")
ind <- sapply(ind, function(x) x[2])
ind <- as.numeric(ind)
table(cell_type, time)
bact <- ifelse(time == 0, "none", "Rv")
colnames(exp_norm) <- sprintf("%s%03d.%s.%d", cell_type, ind, bact, time)
```

Also create an annotation table.

```{r anno-table}
anno_array <- data.frame(ind, bact, time = time, cell_type,
                         stringsAsFactors = FALSE)
head(anno_array)
```

## Visualize distributions

The data do appear to already be normalized.

```{r per-array-boxplot, fig.width = 10}
boxplot(exp_norm, las = 3)
```

## Filter genes

As expected from [Tailleux et al., 2008][Tailleux2008], the file contains data for `r nrow(exp_norm)` Affy probes.

> Comprehensive gene expression profiles of 9 independent healthy donors were generated with high-density oligonucleotide human arrays with 22,283 probe sets, which in total interrogated the expression levels of approximately 18,400 transcripts and variants, including 14,500 well-characterized human genes.

```{r count-probes}
nrow(exp_norm)
head(rownames(exp_norm))
```

However, these were filtered to just 11,262.

> Using unsupervised hierarchical cluster analysis with 11,262 probe sets we identified the differences in gene expression between DCs and Mφs, which readily distinguished the two groups.

The AMDA report explains that it uses the Affy "Present" and "Absent" statistics to obtain the final 11,262 probes.
Since `exprSetSnorm.txt` no longer contains this information, I'll simply take the 11,262 probes with the highest expression level.

```{r filter}
exp_norm_mean <- apply(exp_norm, 1, mean)
stopifnot(sum(is.na(exp_norm_mean)) == 0)
exp_norm_filter <- exp_norm[order(exp_norm_mean, decreasing = TRUE)[1:11262], ]
stopifnot(nrow(exp_norm_filter) == 11262,
          colMeans(exp_norm_filter) > colMeans(exp_norm))
```

As expected, this increased the average expression for the arrays.

```{r per-array-boxplot-filtered, fig.width = 10}
boxplot(exp_norm_filter, las = 3)
```

## PCA

Perform principal components analysis on the filtered data.

```{r run-pca}
pca <- prcomp(x = t(exp_norm_filter), retx = TRUE, center = TRUE, scale. = TRUE)
```

Combine PCA results with annotation.

```{r labels}
pca_dat <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], anno_array,
                      stringsAsFactors = FALSE)
```

Calculate the variance explained.

```{r variance-explained}
variances <- pca$sdev^2
explained <- variances / sum(variances)
```

Plot the first two PCs with the points labeled by cell type and time point.

```{r plot-pca}
pca_plot <- ggplot(pca_dat, aes(x = PC1, y = PC2, col = as.factor(time),
                                shape = cell_type)) +
  geom_point() +
  labs(x = sprintf("PC1 (%.2f%%)", round(explained[1] * 100, 2)),
       y = sprintf("PC2 (%.2f%%)", round(explained[2] * 100, 2)))
pca_plot
```

The first PC divides the samples by cell type.
The second PC divides the samples by time post-infection.

## Convert gene IDs

Convert Affy probe IDs to Ensembl gene IDs.
The array was the Affymetrix Human U133A.

> Labelling of samples and hybridisation to the Human U133A oligonucleotide microarray chips (Affymetrix, Santa Clara, CA) containing 22,283 probe sets were performed according to the manufacturer's protocols.

Set up the connection to Ensembl Biomart.

```{r setup-biomaRt}
ensembl <- useMart(host = "sep2013.archive.ensembl.org",
                   biomart = "ENSEMBL_MART_ENSEMBL",
                   dataset = "hsapiens_gene_ensembl")
# attributePages(ensembl)
# atts <- listAttributes(ensembl, page = "feature_page")
# atts[grep("Affy", atts$description), ]
# filts <- listFilters(ensembl)
# filts[grep("u133a", filts$description), ]
```

Convert the Affy probe IDs to Ensembl gene names.

```{r get-ens-names}
ens_names <- getBM(attributes = c("ensembl_gene_id", "external_gene_id",
                                  "affy_hg_u133a"),
                   filters = "affy_hg_u133a",
                   values = rownames(exp_norm_filter),
                   mart = ensembl)
head(ens_names)
```

Merge the Ensembl gene names into the expression matrix.

```{r merge}
exp_norm_ens <- merge(exp_norm_filter, ens_names, by.x = "row.names", by.y = "affy_hg_u133a",
                      all.x = TRUE)
dim(exp_norm_ens)
```

There are probes which have no Ensembl ID.

```{r no-ens-id}
no_ens_id <- is.na(exp_norm_ens$ensembl_gene_id)
sum(no_ens_id)
head(exp_norm_ens[no_ens_id, c(1:3, 73:75)])
```

Remove them.

```{r remove-na-id}
exp_norm_ens <- exp_norm_ens[!no_ens_id, ]
```

There are also be probes that target the same gene.

```{r duplicate-probes}
multi_probes <- duplicated(exp_norm_ens$ensembl_gene_id)
sum(multi_probes)
```

Take the mean expression across the probes for each gene.

```{r mean-exp}
exp_norm_mean <- exp_norm_ens %>%
  group_by(ensembl_gene_id, external_gene_id) %>%
  summarise_each(funs(mean), -Row.names) %>%
  arrange(ensembl_gene_id, external_gene_id)
head(exp_norm_mean)
```

Confirm that this worked as expected.

```{r confirm-summary}
ex_multi_gene <- exp_norm_ens[which(multi_probes)[1], "ensembl_gene_id"]
exp_norm_ens %>% filter(ensembl_gene_id == ex_multi_gene) %>%
  select(-starts_with("D", ignore.case = FALSE))
stopifnot(exp_norm_mean[exp_norm_mean$ensembl_gene_id == ex_multi_gene, "M001.none.0"] ==
          mean(exp_norm_ens[exp_norm_ens$ensembl_gene_id == ex_multi_gene, "M001.none.0"]))
```

There are no more duplicated probes.

```{r no-dups}
stopifnot(sum(duplicated(exp_norm_mean$ensembl_gene_id)) == 0)
```

## Merge the two data sets

```{r input-counts}
dat_cpm <- read.table("../data/table-s1.txt", header = TRUE, sep = "\t",
                      stringsAsFactors = FALSE)
# Fix Rv+
colnames(dat_cpm) <- gsub("\\.\\.", "\\+\\.", colnames(dat_cpm))
dim(dat_cpm)
```

Merge on the Ensembl gene names.

```{r merge-two-data-sets}
dat_merged <- merge(exp_norm_mean, dat_cpm, by.x = "ensembl_gene_id", by.y = "id")
dim(dat_merged)
```

Confirm that gene names match from both data sets.

```{r check-gene-names}
stopifnot(dat_merged$external_gene_id == dat_merged$name)
```

There are `r nrow(dat_merged)` genes that are able to be compared between the two experiments.

Clean up the expression data to only contain data on macrophages infected with MTB.

```{r clean}
dat_mac <- dat_merged
rownames(dat_mac) <- dat_mac$ensembl_gene_id
dat_mac <- dat_mac %>%
  select(which(grepl("Rv\\.", colnames(dat_mac)) | grepl("none", colnames(dat_mac)))) %>%
  select(starts_with("M", ignore.case = FALSE))
dim(dat_mac)
```

Load the annotation data from the RNA-seq study and select only the MTB infection data.

```{r input-metadata}
anno_seq <- read.table("../data/annotation.txt", header = TRUE,
                       stringsAsFactors = FALSE)
anno_seq <- filter(anno_seq, bact %in% c("Rv", "none"))
```

Combine the annotations.

```{r combine-anno}
anno_seq$study <- "blischak2015"
anno_array$study <- "tailleux2008"
anno <- rbind(anno_array[anno_array$cell_type == "M",
                         c("ind", "bact", "time", "study")],
              anno_seq[, c("ind", "bact", "time", "study")])
stopifnot(anno$bact %in% c("Rv", "none"),
          nrow(anno) == ncol(dat_mac))
head(anno)
```

## PCA to compare studies

Run PCA on the merged macrophage data.

```{r run-pca-macrophage}
pca_mac <- prcomp(x = t(dat_mac), retx = TRUE, center = TRUE, scale. = TRUE)
```

```{r labels-mac}
pca_mac_dat <- data.frame(pca_mac$x, anno, stringsAsFactors = FALSE)
```

```{r variance-explained-mac}
variances_mac <- pca_mac$sdev^2
explained_mac <- variances_mac / sum(variances_mac)
```

Plotting the first two PCs identifies the expected batch effect.

```{r plot-pca-mac-batch}
pca_mac_batch <- ggplot(pca_mac_dat, aes(x = PC1, y = PC2, col = bact,
                                         shape = study)) +
  geom_point() +
  geom_text(aes(label = time), hjust = -.5, vjust = 0, size = 3) +
  labs(title = "Batch effect between studies",
       x = sprintf("PC1 (%.2f%%)", round(explained_mac[1] * 100, 2)),
       y = sprintf("PC2 (%.2f%%)", round(explained_mac[2] * 100, 2)))
pca_mac_batch
```

Plotting PC2 versus PC3 separates the samples by their treatment.

```{r plot-pca-mac-treatment}
pca_mac_treat <- pca_mac_batch %+% aes(x = PC2, y = PC3) +
  labs(title = "Samples separated by treatment",
       x = sprintf("PC2 (%.2f%%)", round(explained_mac[2] * 100, 2)),
       y = sprintf("PC3 (%.2f%%)", round(explained_mac[3] * 100, 2)))
pca_mac_treat
```

Use a polygon to highlights the samples that demonstrate a strong response to infection.

```{r plot-pca-mac-treatment-poly}
# coordinates of polygon go counter clockwise starting from bottom left
polygons <- data.frame(id = "response-to-infection",
                       x = c(-07, -07, +45, +45),
                       y = c(-58, +10, +10, -58))
pca_mac_treat_poly <- pca_mac_treat + geom_polygon(mapping = aes(x = x, y = y, group = id,
                                           col = NA, shape = NA),
                             alpha = 0, data = polygons)
pca_mac_treat_poly
```

```{r}
pdf("../figs/fig-S10-tailleux2008.pdf", width = 8, height = 7,
    useDingbats = FALSE)
grid.arrange(pca_mac_batch, pca_mac_treat_poly)
dev.off()
```


```{r info}
sessionInfo()
```
